One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-08-06 papers/1_dmatpol/paper-submit.tex
Revisiom papers added

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix 
susceptibilities (DMS) interacting with the spatial distribution of electric field.
The analytical approximate expressions are derived at the Hartree\hyp{}Fock theory, 
which serves as a basis for construction of generalized model 
that can be used for an arbitrary form of wavefunction and any type of one\hyp{}particle density matrix
as long as the latter are available by means of quantum chemistry methods. 
It is shown that it is possible to accurately predict 
the changes in the one\hyp{}electron ground\hyp{}state density of water molecule 
in spatially uniform electric field, as well as in spatially non\hyp{}uniform 
electric field distribution generated by point charges. When both linear 
and quadratic terms with respect to the electric field are accounted for,
the electric field\hyp{}induced polarization energies, dipole moments 
and quadrupole moments are quantitatively described by the present theory 
in electric fields ranging from weak to very strong (0.001 -- 0.07 a.u.).
It is believed that the proposed model could open new routes in quantum chemistry 
for fast and efficient calculations of molecular properties in condensed phases.
\end{abstract}

@@

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories 
when large and complex chemical systems such as
solutions and biomolecules come into play.\cite{Tomasi.Mennucci.Cammi.ChemRev.2005}
Due to this reason, it has been of particular importance nowadays 
to develop bridging theoretical techniques that link
extremely accurate but localized pieces of information 
with certain knowledge of an atomistic organisation and intermolecular interaction
on a relatively large spatial scales, typically ranging beyond few nanometers.

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry which allows insightful theoretical studies
of quantum mechanical effects in biological systems because the costly quantum mechanical calculations
can be combined with lower\hyp{}level and much less expensive
classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
Understanding physics that rules intermolecular interactions\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
helped to develop
a wide family of force fields, and served as a basis for embedding Hamiltonian
within an effective environment and fragmentation methods.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}

However, even with the modern advances in computational chemistry, 
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,
Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017} 
Interface between QM and MM levels is usually realized by electrostatic embedding
and only rarely the Pauli exclusion principle and dispersion are additionally included.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
due to the complicated nature of intermolecular interactions that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
One of the most frequently used, efficient and reasonably accurate \emph{ab initio} fragmentation methods,
the effective fragment potential (EFP) methods, are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF)\cite{Roothaan.RevModPhys.1951,Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.

Most of the efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{}based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wider scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,\cite{Holas.March.PhysRevA.1991}
though obtaining the accurate electron density distribution has been an undersolved issue for most of chemical systems
in condensed phases.
Given that fact, it would be incredibly insightful 
to establish a first\hyp{}principles connection between the one\hyp{}particle
density and the perturbation in the environment through introduction of
\emph{density matrix susceptibility} (DMS) tensors. 
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%
\begin{equation*} \label{e:initial-model}
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential which in principle, is applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
because the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.

The simplest type of perturbation, yet the most relevant for chemistry, 
is the electrostatic perturbation.
It is worth mentioning that, at the moment, there is no simple \emph{ab initio}
method to compute polarization\hyp{}induced electron deformation densities 
without considering the entire molecular 
aggregate by using full quantum chemistry calculations.\cite{Horn.Head-Gordon.JCP.2015,
Mandado.Hermida-Ramon.JCTC.2011}
In this work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion 
or dispersion,\cite{Mandado.Hermida-Ramon.JCTC.2011}
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.

@@

%
\begin{table*}%[]
\caption{{\bf Performance of the DMS tensors of water molecule
in uniform and non\hyp{}uniform electric field$^{a,b}$}
}
\label{t:results}
\begin{ruledtabular}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cddddddcddddcdd}
%\hline\hline
\toprule
\multicolumn{1}{l}{} & \multicolumn{6}{c}{\textbf{Ab Initio DMS}} &  & \multicolumn{4}{c}{\textbf{Generalized DMS}}
                                                                  &  & \multicolumn{2}{c}{\textbf{Distributed}} \\
\cline{2-8}
\cline{9-12}
%\cline{14-15}
\textbf{}         
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^c$}
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^d$}
& \multicolumn{2}{c}{$\BM{R_{\rm max}=2}$$^d$}
&
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^e$} 
& \multicolumn{2}{c}{$\BM{R_{\rm max}=2}$$^e$} 
&  
& \multicolumn{2}{c}{\textbf{Polarizability}}        \\ \hline
\multicolumn{15}{c}{} \\ 
\multicolumn{15}{c}{\textbf{Uniform Electric Field}} \\ 
\multicolumn{15}{c}{} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSE (kcal/mol)}}              \\ %\hline
   & 4.041 & (n.d) & 4.823 & (n.d.) & 0.231 & (98.40) && 5.0121 & (n.d.) & 0.173 & (99.10) && 0.112 & (99.63) \\ 
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSD (mD)}}                    \\ %\hline
   & 30.46 & (97.91) & 30.46 & (97.91) & 18.02 & (99.27) && 24.83 & (98.61) & 7.13 & (99.89) && 30.46 & (97.91) \\ 
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSQ (D\AA $\times 10^{-3}$)}} \\ %\hline
   & 47.89 & (69.17) & 39.24 & (79.30) &  8.56 & (99.02) && 37.63 & (80.96) & 3.94 & (99.79) && 50.66 & (65.49) \\ \hline
\multicolumn{15}{c}{} \\ 
\multicolumn{15}{c}{\textbf{Non-Uniform Electric Field}$^{f}$} \\ 
\multicolumn{15}{c}{} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSE (kcal/mol)}}   \\ 
W & 0.067 & (n.d.)     & - & (-) & - & (-) && 0.076 & (-)     & 0.003 & (99.45) && 0.004 & (98.64)\\ 
M & 1.072 & (n.d.)     & - & (-) & - & (-) && 1.236 & (-)     & 0.057 & (99.08) && 0.070 & (98.58)\\ 
S & 4.220 & (n.d.)     & - & (-) & - & (-) && 5.147 & (-)     & 0.402 & (97.25) && 0.332 & (98.13) \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSD (mD)}}  \\ 
W &  3.36 & (98.71) & - & (-) & -  & (-) &&  1.03 & (99.88) &  0.94 & (99.90) &&  3.36 & (98.71) \\ 
M & 14.72 & (98.50) & - & (-) & -  & (-) &&  7.72 & (99.59) &  4.55 & (99.86) && 14.72 & (98.50)\\ 
S & 44.80 & (96.92) & - & (-) & -  & (-) && 33.06 & (98.33) & 15.96 & (99.61) && 44.80 & (96.92) \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSQ (D\AA $\times 10^{-3}$)}} \\ 
W &  6.22 & (58.72) & - & (-) & - & (-) &&  1.59 & (97.32) &  1.42 & (97.86) &&  5.45 & (68.29) \\ 
M & 28.22 & (53.90) & - & (-) & - & (-) && 10.45 & (93.68) &  5.68 & (98.13) && 23.94 & (66.83) \\ 
S & 78.07 & (40.53) & - & (-) & - & (-) && 43.33 & (81.68) & 14.40 & (97.98) && 64.33 & (59.62) \\ 
\bottomrule
\end{tabular}%
}
\end{ruledtabular}
%
\footnotesize{\flushleft{
$^a$ $R$-squared coefficients are shown in parentheses. `$n.d.$' stands for `not determined'. 
$^b$ Ranks of the density matrix susceptibility models are denoted as $R_{\rm max}$ 
where $R_{\rm max}$ is the maximum power in the expansion from Eq.~\eqref{e:final-model.General}.
$^c$ \emph{Ab initio} HF model, Eq.~\eqref{e:final-model.HF}.
$^d$ \emph{Ab initio} finite\hyp{}difference model, Eq.~\eqref{e:dD-Taylor}.
$^e$ Linear\hyp{}regression generalized model, Eq.~\eqref{e:final-model.General}.
$^f$ `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields 
that are generated by sets of point charges, respectively (c.f. Table~\ref{t:fields}).\\
}}
\end{table*}

@@

Similarly as in uniform electric field, 
the \emph{ab initio} DMS model from Eq.~\eqref{e:final-model.HF} 
as well as
the linear generalized DMS model, are insufficient to correctly describe the polarization energy 
in non\hyp{}uniform electric field,
although the correlation with benchmark is still very clear with a slope of $\sim$2
(Figure~\ref{f:fig-2}A). 
Again, induced dipole moments are reproduced very well also
in strong electric fields (RMSD of 44.8~mD and lesser, $R^2>96\%$).
Induced quadrupole moments are well described only by the generalized DMS model.
For example, in medium electric fields RMSQ is 10.4$\times 10^{-3}$D\AA{ }with $R^2=93.7\%$
whereas the \emph{ab initio} DMS model yealds RMSQ of 28.2$\times 10^{-3}$D\AA{ }($R^2=53.9\%$).

>>>>

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix 
susceptibilities (DMS) interacting with the spatial distribution of electric field.
The analytical approximate expressions are derived at the Hartree\hyp{}Fock theory, 
which serves as a basis for construction of generalized model 
that is designed for an arbitrary form of wavefunction and any type of one\hyp{}particle density matrix.
It is shown that it is possible to accurately predict 
the changes in the one\hyp{}electron ground\hyp{}state density of water molecule 
in spatially uniform electric field, as well as in spatially non\hyp{}uniform 
electric field distribution generated by point charges. When both linear 
and quadratic terms with respect to the electric field are accounted for,
the electric field\hyp{}induced polarization energies, dipole moments 
and quadrupole moments are quantitatively described by the present theory 
in electric fields ranging from weak to very strong (0.001 -- 0.07 a.u.).
It is believed that the proposed model could open new routes in quantum chemistry 
for fast and efficient calculations of molecular properties in condensed phases.
\end{abstract}

@@

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories when large and complex 
chemical systems such as solutions and biomolecules come into play.\cite{
Xu.Guidez.Bertoni.Gordon.JCP.2018,Tomasi.Mennucci.Cammi.ChemRev.2005}
Due to this reason, it has been of particular importance nowadays to develop bridging theoretical techniques 
that link extremely accurate but localized pieces of information with certain knowledge of an atomistic 
organisation and intermolecular interaction on a relatively large spatial scales, typically ranging beyond 
few nanometers.

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM) technique 
was a revolution in computational chemistry which allows insightful theoretical studies of quantum mechanical
effects in biological systems because costly quantum mechanical calculations can be combined with lower\hyp{}level 
and much less expensive classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
Fruitful incorporation of accurate \emph{ab initio} methods with extensive experimental and semi\hyp{}empirical
refinings led to development of a wide variety of force fields,\cite{Demerdash.Yap.T-Head-Gordon.AnnuRevPhysChem.2014}
whereas sophisticated partitioning of Hamiltonian or electron density enabled numerous schemes for efficient fragmentation
within a QM level.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
%Gao.Truhlar.Wang.Mazack.Loffler.Provorse.Rehak.ACR.2014}

However, even with the modern advances in computational chemistry, accurate description of phenomena due 
to complex heterogeneous molecular environments remains a challenge in the following areas: (i) describing 
the electronically excited states (electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017,
Xu.Blasiak.Cho.Layfield.Londergan.JPCL.2018} 
Interface between QM and MM levels is usually realized by electrostatic embedding and only rarely 
the Pauli exclusion principle and dispersion are additionally included.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
due to complicated nature of intermolecular interactions that theoretically cannot be solely envisioned 
as a sum of separate contributions coming from electrostatic, dispersion and Pauli 
effects.\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
For example,
one of the most frequently used, efficient and reasonably accurate \emph{ab initio} fragmentation methods, 
the effective fragment potential (EFP) methods, are limited to the Hartree\hyp{}Fock 
(HF)\cite{Roothaan.RevModPhys.1951,Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Hohenberg.Kohn.PhysRev.1964,Kohn.Sham.PhysRev.1965,Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.

Most of the efforts to theoretically treat extended molecular aggregates are rooted in energy\hyp{}based 
philosophy. However, in the past two decade there has been considerable interest in density\hyp{}based approaches
of a wider scope and aims.\cite{Yang.Lee.JCP.1995,
Mukamel.Berman.JCP.2003,
Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Elliott.Burke.Cohen.Wasserman.PhysRevA.2010,
Mandado.Hermida-Ramon.JCTC.2011,Sun.Chan.ACR.2016,Hedegard.Reiher.JCTC.2016} 
In particular, the importance of one\hyp{}electron densities in chemistry is well known and acknowledged 
in the community,\cite{Holas.March.PhysRevA.1991,Yang.PhysRevLett.1991,Yang.Lee.JCP.1995} 
though obtaining the accurate electron density distribution 
has been an undersolved issue for most of chemical systems in condensed phases. Given that fact, it would be 
incredibly insightful to establish a first\hyp{}principles connection between the one\hyp{}particle density 
and the perturbation in the environment through introduction of \emph{density matrix susceptibility} (DMS) tensors. 
The main postulate of the present work is that the associated change in the one\hyp{}particle density matrix, 
$\delta {\bf D}$, can be represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$, evaluated at certain site $i$,
%
\begin{equation*} \label{e:initial-model}
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective one\hyp{}electron fragment potential which in principle, 
is applicable to an arbitrary level of theory, as long as the one\hyp{}particle densities are available by means of
quantum chemistry methods. It also means that such a technique would dramatically reduce costs of quantum chemistry 
calculations because the task of performing full computation on a molecular aggregate would reduce to much simpler 
matrix operations.

The simplest type of perturbation, yet the most relevant for chemistry, is the electrostatic perturbation.
It is worth mentioning that, at the moment, there is no simple \emph{ab initio} method to compute 
polarization\hyp{}induced electron deformation densities without considering the entire molecular 
aggregate by using full quantum chemistry calculations.\cite{
Ma.Politzer.JCP.2004,
Horn.Head-Gordon.JCP.2015,
Mandado.Hermida-Ramon.JCTC.2011}
In this work, we explore the susceptibility of the molecule's electronic density
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion 
or dispersion,\cite{Mandado.Hermida-Ramon.JCTC.2011}
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.

@@

%
\begin{table*}%[]
\caption{{\bf Performance of the DMS tensors of water molecule
in uniform and non\hyp{}uniform electric field$^{a,b}$}
}
\label{t:results}
\begin{ruledtabular}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cddddddcddddcdd}
%\hline\hline
\toprule
\multicolumn{1}{l}{} & \multicolumn{6}{c}{\textbf{Ab Initio DMS}} &  & \multicolumn{4}{c}{\textbf{Generalized DMS}}
                                                                  &  & \multicolumn{2}{c}{\textbf{Distributed}} \\
\cline{2-8}
\cline{9-12}
%\cline{14-15}
\textbf{}         
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^c$}
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^d$}
& \multicolumn{2}{c}{$\BM{R_{\rm max}=2}$$^d$}
&
& \multicolumn{2}{c}{$\BM{R_{\rm max}=1}$$^e$} 
& \multicolumn{2}{c}{$\BM{R_{\rm max}=2}$$^e$} 
&  
& \multicolumn{2}{c}{\textbf{Polarizability}}        \\ \hline
\multicolumn{15}{c}{} \\ 
\multicolumn{15}{c}{\textbf{Uniform Electric Field}} \\ 
\multicolumn{15}{c}{} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSE (kcal/mol)}}              \\ %\hline
   & 4.041 & (n.d) & 4.823 & (n.d.) & 0.231 & (98.40) && 5.0121 & (n.d.) & 0.173 & (99.10) && 0.112 & (99.63) \\ 
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSD (mD)}}                    \\ %\hline
   & 30.46 & (97.91) & 30.46 & (97.91) & 18.02 & (99.27) && 24.83 & (98.61) & 7.13 & (99.89) && 30.46 & (97.91) \\ 
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSQ (D\AA $\times 10^{-3}$)}} \\ %\hline
   & 47.89 & (69.17) & 39.24 & (79.30) &  8.56 & (99.02) && 37.63 & (80.96) & 3.94 & (99.79) && 50.66 & (65.49) \\ \hline
\multicolumn{15}{c}{} \\ 
\multicolumn{15}{c}{\textbf{Non-Uniform Electric Field}$^{f}$} \\ 
\multicolumn{15}{c}{} \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSE (kcal/mol)}}   \\ 
W & 0.067 & (n.d.)  & - &  -  & - &  -  && 0.076 & (n.d.)  & 0.003 & (99.45) && 0.004 & (98.64) \\ 
M & 1.072 & (n.d.)  & - &  -  & - &  -  && 1.236 & (n.d.)  & 0.057 & (99.08) && 0.070 & (98.58) \\ 
S & 4.220 & (n.d.)  & - &  -  & - &  -  && 5.147 & (n.d.)  & 0.402 & (97.25) && 0.332 & (98.13) \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSD (mD)}}  \\ 
W &  3.36 & (98.71) & - &  -  & - &  -  &&  1.03 & (99.88) &  0.94 & (99.90) &&  3.36 & (98.71) \\ 
M & 14.72 & (98.50) & - &  -  & - &  -  &&  7.72 & (99.59) &  4.55 & (99.86) && 14.72 & (98.50) \\ 
S & 44.80 & (96.92) & - &  -  & - &  -  && 33.06 & (98.33) & 15.96 & (99.61) && 44.80 & (96.92) \\ \hline
\multicolumn{1}{l}{} & \multicolumn{14}{c}{\textbf{RMSQ (D\AA $\times 10^{-3}$)}} \\ 
W &  6.22 & (58.72) & - &  -  & - &  -  &&  1.59 & (97.32) &  1.42 & (97.86) &&  5.45 & (68.29) \\ 
M & 28.22 & (53.90) & - &  -  & - &  -  && 10.45 & (93.68) &  5.68 & (98.13) && 23.94 & (66.83) \\ 
S & 78.07 & (40.53) & - &  -  & - &  -  && 43.33 & (81.68) & 14.40 & (97.98) && 64.33 & (59.62) \\ 
\bottomrule
\end{tabular}%
}
\end{ruledtabular}
%
\footnotesize{\flushleft{
$^a$ $R$-squared coefficients are shown in parentheses. `$n.d.$' stands for `not determined'. 
$^b$ Ranks of the density matrix susceptibility models are denoted as $R_{\rm max}$ 
where $R_{\rm max}$ is the maximum power in the expansion from Eq.~\eqref{e:final-model.General}.
$^c$ \emph{Ab initio} HF model, Eq.~\eqref{e:final-model.HF}.
$^d$ \emph{Ab initio} finite\hyp{}difference model, Eq.~\eqref{e:dD-Taylor}.
$^e$ Linear\hyp{}regression generalized model, Eq.~\eqref{e:final-model.General}.
$^f$ `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields 
that are generated by sets of point charges, respectively (c.f. Table~\ref{t:fields}).\\
}}
\end{table*}

@@

Similarly as in uniform electric field, 
the \emph{ab initio} DMS model from Eq.~\eqref{e:final-model.HF} 
as well as
the generalized linear DMS model, are insufficient to correctly describe the polarization energy 
in non\hyp{}uniform electric field,
although the correlation with benchmark is still very clear with a slope of $\sim$2
(Figure~\ref{f:fig-2}A). 
Again, induced dipole moments are reproduced very well also
in strong electric fields (RMSD of 44.8~mD and lesser, $R^2>96\%$).
Induced quadrupole moments are well described only by the generalized DMS model.
For example, in medium electric fields RMSQ is 10.4$\times 10^{-3}$D\AA{ }with $R^2=93.7\%$
whereas the \emph{ab initio} DMS model yealds RMSQ of 28.2$\times 10^{-3}$D\AA{ }($R^2=53.9\%$).
